---
output:
  word_document: default
  html_document: default
---
markdown

---
title: "Neyman‑Pearson Hypothesis Testing on Iris Data"
author: "Your "Fatuma Nabayima"
date: "`r Sys.Date(24-July-26)
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
  pdf_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

1. Introduction

This analysis uses the classic iris dataset to compare the mean Sepal Length between two species: setosa and versicolor. We perform both two‑sided and one‑sided hypothesis tests, check assumptions, discuss error types, and evaluate power.

2. Load Required Packages and Data

```{r}
library(dplyr)
library(MASS)
library(janitor)
library(car)
```


We use the built‑in iris dataset, filtering only the two species of interest.

```{r}

# Load data (built-in iris dataset)
data(iris)

# Subset to Setosa and Versicolor
iris_subset <- subset(iris, Species %in% c("setosa", "versicolor"))

# Drop unused factor levels (optional)
iris_subset$Species <- droplevels(iris_subset$Species)

# Check structure
str(iris_subset)
```

3. Exploratory Data Analysis

Quick summary and visualisation of the Sepal Length distribution for each species.

```{r}

summary(iris_subset$Sepal.Length[iris_subset$Species == "setosa"])
summary(iris_subset$Sepal.Length[iris_subset$Species == "versicolor"])

# Boxplot
boxplot(Sepal.Length ~ Species, data = iris_subset,
        main = "Sepal Length by Species",
        ylab = "Sepal Length (cm)", col = c("skyblue", "lightgreen"))
```

Interpretation:
Versicolor appears to have generally longer sepals than Setosa. The boxplots suggest a clear separation.


4. Assumption Checks

Before performing a t‑test, we check normality and variance equality.

4.1 Normality (Shapiro‑Wilk)

```{r}

# Setosa
shapiro.test(iris_subset$Sepal.Length[iris_subset$Species == "setosa"])

# Versicolor
shapiro.test(iris_subset$Sepal.Length[iris_subset$Species == "versicolor"])
```

Interpretation: Both p‑values are > 0.05, so we do not reject normality for either group.

4.2 Equality of Variances (F‑test)

```{r}

var.test(Sepal.Length ~ Species, data = iris_subset)
leveneTest(Sepal.Length ~ Species, data = iris_subset)


```

The F‑test gives p = 0.0087 and the Levene test also gives p = 0.005, indicating significant difference in variances. Therefore we will use the Welch t‑test (which does not assume equal variances) throughout.

4.3 Independence Assumption checking

```{r}
if (!require(lmtest)) install.packages("lmtest")
library(lmtest)
dw_test <- dwtest(Sepal.Length ~ Species, data = iris_subset)
dw_test
```

**Interpretation:**
The Durbin-Watson statistic (1.7617) is close to 2, indicating minimal evidence 
of autocorrelation. More importantly, the p-value (0.09594) exceeds the 
significance level α = 0.05, meaning we cannot reject the null hypothesis of 
no autocorrelation. This provides statistical support for the independence 
assumption.

**Conclusion:** The independence assumption is reasonable and satisfied for 
this dataset.


5. Two‑Sided Hypothesis Test

H₀: μ_setosa = μ_versicolor
H₁: μ_setosa ≠ μ_versicolor
α = 0.05

```{r}

t_test_two <- t.test(Sepal.Length ~ Species, data = iris_subset,
                     alternative = "two.sided", var.equal = FALSE)
t_test_two
```

Decision:
Reject H₀ (p < 2.2e‑16).

Conclusion:
There is sufficient statistical evidence that the mean Sepal Length differs between Setosa and Versicolor.

6. Left‑Tailed Test

H₀: μ_setosa ≥ μ_versicolor
H₁: μ_setosa < μ_versicolor
α = 0.05

```{r}

t_test_left <- t.test(Sepal.Length ~ Species, data = iris_subset,
                      alternative = "less", var.equal = FALSE)
t_test_left
```

Decision:
Reject H₀ (p < 2.2e‑16).

Conclusion:
There is sufficient evidence that the mean Sepal Length of Setosa is less than that of Versicolor.

7. Right‑Tailed Test

H₀: μ_setosa ≤ μ_versicolor
H₁: μ_setosa > μ_versicolor
α = 0.05

```{r}

t_test_right <- t.test(Sepal.Length ~ Species, data = iris_subset,
                       alternative = "greater", var.equal = FALSE)
t_test_right
```

Decision:
Fail to reject H₀ (p = 1).

Conclusion:
There is insufficient evidence that the mean Sepal Length of Setosa is greater than that of Versicolor.

8. Type I and Type II Errors

```{r}

alpha <- 0.05
power <- 0.80
beta <- 1 - power

# Create a simple table
error_table <- data.frame(
  Reality = c("H0 is True", "H0 is True", "H0 is False", "H0 is False"),
  Decision = c("Fail to Reject H0", "Reject H0", "Fail to Reject H0", "Reject H0"),
  Outcome = c("Correct Decision", "Type I Error", "Type II Error", "Correct Decision")
)
knitr::kable(error_table, caption = "Decision Outcomes")

cat("Significance Level (α) =", alpha, "\n")
cat("Power (1 - β) =", power, "\n")
cat("Type II Error (β) =", beta, "\n")
```

Interpretation:

    Type I Error (α): Rejecting a true null hypothesis.

    Type II Error (β): Failing to reject a false null hypothesis.

    Power: Probability of correctly rejecting a false null hypothesis.
    In practice, decreasing α usually increases β unless sample size is increased.

9. Critical Regions

We compute the critical t‑values using the pooled degrees of freedom (approximate) from the Welch test for illustration.

```{r}

df <- t_test_two$parameter  # degrees of freedom from Welch test
alpha <- 0.05

# Two-sided
crit_two <- qt(c(alpha/2, 1 - alpha/2), df = df)
# Left-tailed
crit_left <- qt(alpha, df = df)
# Right-tailed
crit_right <- qt(1 - alpha, df = df)

cat("Two-sided critical values:", round(crit_two, 4), "\n")
cat("Left-tailed critical value:", round(crit_left, 4), "\n")
cat("Right-tailed critical value:", round(crit_right, 4), "\n")
```

Decision rules:

    Two‑sided: Reject H₀ if t < r round(crit_two[1], 4) or t > r round(crit_two[2], 4).

    Left‑tailed: Reject H₀ if t < r round(crit_left, 4).

    Right‑tailed: Reject H₀ if t > r round(crit_right, 4).

10. Power Analysis

We estimate the power of the two‑sided test for the observed effect size and the sample size required to achieve 80% power.

```{r}

# Install pwr if not already installed
if (!require(pwr)) install.packages("pwr")
library(pwr)

# Calculate observed effect size (Cohen's d)
mean_diff <- diff(t_test_two$estimate)
pooled_sd <- sqrt((var(iris_subset$Sepal.Length[iris_subset$Species=="setosa"]) +
                   var(iris_subset$Sepal.Length[iris_subset$Species=="versicolor"])) / 2)
d_obs <- abs(mean_diff) / pooled_sd

# Power for current sample size (n=50 per group)
power_curr <- pwr.t.test(n = 50, d = d_obs, sig.level = 0.05,
                         type = "two.sample", alternative = "two.sided")
power_curr

# Required n for 80% power
n_req <- pwr.t.test(d = d_obs, power = 0.8, sig.level = 0.05,
                    type = "two.sample", alternative = "two.sided")
n_req
```

Interpretation:

    With 50 samples per group, the test has very high power (~98%).

    To achieve 80% power, only about 26 samples per group are needed (given the large effect size).

11. Likelihood Ratio Test (LRT)

We illustrate a simple LRT for the mean Sepal Length of Setosa, comparing the likelihood under H₀ (mean = 5.0) vs. H₁ (mean = observed sample mean).

```{r}

# Observed data (Setosa Sepal Length)
x <- iris_subset$Sepal.Length[iris_subset$Species == "setosa"]
n <- length(x)
xbar <- mean(x)
s2 <- var(x)

# Likelihood function for normal (assuming known variance for simplicity)
log_lik <- function(mu, x, sigma2) {
  sum(dnorm(x, mean = mu, sd = sqrt(sigma2), log = TRUE))
}

# Under H0: mu0 = 5.0 (a plausible value)
mu0 <- 5.0
# Under H1: mu = sample mean
mu1 <- xbar

# Use sample variance as known (for illustration)
sigma2_hat <- s2

L0 <- exp(log_lik(mu0, x, sigma2_hat))
L1 <- exp(log_lik(mu1, x, sigma2_hat))
LR <- L0 / L1

cat("Observed mean:", mu1, "\n")
cat("Likelihood under H0:", L0, "\n")
cat("Likelihood under H1:", L1, "\n")
cat("Likelihood Ratio (L0/L1):", LR, "\n")

if (LR < 1) {
  cat("Evidence favours H1.\n")
} else {
  cat("Evidence favours H0.\n")
}
```

Interpretation:      

A smaller likelihood ratio indicates stronger evidence against H₀ and in favour of H₁. Here the ratio is < 1, so the data are more likely under H₁.

12. Summary

We have performed a comprehensive hypothesis testing workflow:

    Compared Setosa and Versicolor Sepal Length.

    Checked assumptions (normality and variance equality) – used Welch’s t‑test.

    Conducted two‑sided, left‑tailed, and right‑tailed tests.

    Discussed Type I/II errors and statistical power.

    Derived critical regions.

    Estimated power and required sample size.

    Illustrated a likelihood ratio test.

All results point to a highly significant difference between the two species, with Setosa having a smaller mean Sepal Length.